Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2019-10-28 ms/manuscript.Rmd
Update MS
Output eps image format for MS submission

Standardize 'pinna no.'

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)(c, d)). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In the standard PCA, size-related traits tended to load on PC1, with SLA, pinna number, and degree of dissection on PC2. In the phylogenetic PCA, SLA was strongly loaded on PC1, while dissection, pinna, and other size-related traits were mostly loaded on PC2.

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under BM (\tab \@ref(tab:phylosig-binary)). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)). Similar results were obtained using the "strict" dataset (\tab S4).

@@

Community-weighted mean values (CWMs) were smaller for epiphytic than terrestrial ferns for size-related traits (frond length, frond width, stipe length, and rhizome diameter) and pinna number (\tval-test, all \pval < 0.05; \fig \@ref(fig:cwm-div)). Specific leaf area decreased with elevation, and was slightly lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Frond length and width and rhizome diameter decreased slightly with elevation for terrestrial communities, while increasing slightly for epiphytic communities (\fig \@ref(fig:cwm-div)).

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (\tab \@ref(tab:pics)), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (\fig S2).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird’s nest ferns (e.g., *Asplenium nidus*), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird’s nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale has been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA) (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; @Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes (\fig \@ref(fig:cwm-div)).
This may reflect adaptations to specific niches that are not captured in the binary categorization of habitat commonly applied in ferns. 
For example, the following species ranked among those with the five lowest SLA values (i.e., thick, well-protected leaves) on Moorea: *Humata anderssonii* (a twig epiphyte), *Acrostichum aureum* (a terrestrial mangrove specialist), and *Paesia divaticatissima* (a terrestrial species occurring on high-elevation ridges) (Table S2).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

@@

**\fig \@ref(fig:traits-used-image)**: Examples of traits used in this study. **(a)** Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinna pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. **(b)** Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars 1 mm except for glands, which is 0.1 mm. Photographs by J. H. Nitta.

@@

```{r climate, fig.height = 4, fig.width = 7, fig.cap = "Microclimatic variables."}
climate_plot
ggsave(plot = climate_plot, filename = here::here("ms/Fig_2.pdf"), height = 4, width = 7, units = "in")
```

@@

```{r pca, fig.height = 6, fig.width = 6, fig.cap = "Trait PCA."}
pca_plot
ggsave(plot = pca_plot, filename = here::here("ms/Fig_3.pdf"), height = 6, width = 6, units = "in")
```

@@

```{r tree, fig.height = 8, fig.width = 6, fig.cap = "Traits on tree."}
traits_with_tree
ggsave(plot = traits_with_tree, filename = here::here("ms/Fig_4.pdf"), height = 8, width = 6, units = "in")
```

@@

```{r heatmap, fig.height = 4, fig.width = 6, fig.cap = "Variable importance scores."}
importance_heatmap
ggsave(plot = importance_heatmap, filename = here::here("ms/Fig_5.pdf"), height = 4, width = 6, units = "in")
```

@@

```{r cwm-div, fig.height = 4, fig.width = 7, fig.cap = "Community-weighted means."}
combined_cwm_plots
ggsave(plot = combined_cwm_plots, filename = here::here("ms/Fig_6.pdf"), height = 4, width = 7, units = "in")
```

@@

```{r comm-div, fig.height = 8, fig.width = 4, fig.cap = "Community diversity metrics."}
combined_comm_div_plots
ggsave(plot = combined_comm_div_plots, filename = here::here("ms/Fig_7.pdf"), height = 8, width = 4, units = "in")
```

>>>>

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)(c, d)). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In the standard PCA, size-related traits tended to load on PC1, with SLA, number of pinnae pairs, and degree of dissection on PC2. In the phylogenetic PCA, SLA was strongly loaded on PC1, while dissection, pinnae, and other size-related traits were mostly loaded on PC2.

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinnae pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under BM (\tab \@ref(tab:phylosig-binary)). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)). Similar results were obtained using the "strict" dataset (\tab S4).

@@

Community-weighted mean values (CWMs) were smaller for epiphytic than terrestrial ferns for size-related traits (frond length, frond width, stipe length, and rhizome diameter) and number of pinnae pairs (\tval-test, all \pval < 0.05; \fig \@ref(fig:cwm-div)). Specific leaf area decreased with elevation, and was slightly lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Frond length and width and rhizome diameter decreased slightly with elevation for terrestrial communities, while increasing slightly for epiphytic communities (\fig \@ref(fig:cwm-div)).

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (\tab \@ref(tab:pics)), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (\fig S2).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird’s nest ferns (e.g., *Asplenium nidus* L.), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird’s nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale has been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA) (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; @Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes (\fig \@ref(fig:cwm-div)).
This may reflect adaptations to specific niches that are not captured in the binary categorization of habitat commonly applied in ferns. 
For example, the following species ranked among those with the five lowest SLA values (i.e., thick, well-protected leaves) on Moorea: *Humata anderssonii* (Mett.) C.Chr. (a twig epiphyte), *Acrostichum aureum* L. (a terrestrial mangrove specialist), and *Paesia divaticatissima* (Dryand.) Copel. (a terrestrial species occurring on high-elevation ridges) (Table S2).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

@@

**\fig \@ref(fig:traits-used-image)**: Examples of traits used in this study. **(a)** Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinnae pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. **(b)** Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars 1 mm except for glands, which is 0.1 mm. Photographs by J. H. Nitta.

@@

```{r climate, fig.height = 4, fig.width = 7, fig.cap = "Microclimatic variables."}
climate_plot
ggsave(plot = climate_plot, filename = here::here("ms/Fig_2.eps"), height = 4, width = 7, units = "in")
```

@@

```{r pca, fig.height = 6, fig.width = 6, fig.cap = "Trait PCA."}
pca_plot
ggsave(plot = pca_plot, filename = here::here("ms/Fig_3.eps"), height = 6, width = 6, units = "in")
```

@@

```{r tree, fig.height = 8, fig.width = 6, fig.cap = "Traits on tree."}
traits_with_tree
ggsave(plot = traits_with_tree, filename = here::here("ms/Fig_4.eps"), height = 8, width = 6, units = "in")
```

@@

```{r heatmap, fig.height = 4, fig.width = 6, fig.cap = "Variable importance scores."}
importance_heatmap
ggsave(plot = importance_heatmap, filename = here::here("ms/Fig_5.eps"), height = 4, width = 6, units = "in")
```

@@

```{r cwm-div, fig.height = 4, fig.width = 7, fig.cap = "Community-weighted means."}
combined_cwm_plots
ggsave(plot = combined_cwm_plots, filename = here::here("ms/Fig_6.eps"), height = 4, width = 7, units = "in")
```

@@

```{r comm-div, fig.height = 8, fig.width = 4, fig.cap = "Community diversity metrics."}
combined_comm_div_plots
ggsave(plot = combined_comm_div_plots, filename = here::here("ms/Fig_7.eps"), height = 8, width = 4, units = "in")
```
